---
title: "Troubleshooting bad rasters from DL predictions"
format: html
---


```{r}
#| label: setup

library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(stringr)
library(here)
here()

```


Current list of bad rasters

- cluster 82 huc 042701010203
- cluster 218 huc 041300030301
- cluster 218 huc 041300030206
- cluster 218 huc 041402010101
- cluster 11 huc 042900030101
- cluster 11 huc 042900030601
- cluster 225 huc 042900050201
- cluster 22 huc 043002010302

```{r}
bad_list <- list("042701010203", "041300030301", "041300030206", "041402010101", "042900030101", "042900030601", "042900050201", "043002010302")
```


To check plot the raster stacks with a subset of the names

- DEM, slope, curve, twi, CHM, EVI, r (red naip), pct_below_1m 

```{r}
#| eval: false
rast("Data/HUC_Raster_Stacks/HUC_DL_Stacks/cluster_82_huc_042701010203_stack.tif", lyrs = c("DEM", "slope_local", "Geomorph_local", "meanc_local", "twi", "CHM", "EVI", "r", "pct_below_1m" )) |> plot()
```

Function version: 
```{r}
plot_fun <- function(huc){
  l <- list.files(here("Data/HUC_Raster_Stacks/HUC_DL_Stacks/"), full.names = TRUE)
  fn <- l[grepl(huc, l)]
  cluster <- str_extract(fn, "(?<=cluster_)\\d+(?=_)")
  # fn <- paste0("Data/HUC_Raster_Stacks/HUC_DL_Stacks/cluster_", cluster, "_huc_", huc, "_stack.tif")
  r <- rast(here(fn), lyrs = c("DEM", "twi", "CHM", "EVI", "r", "pct_below_1m" ))
  plot(r, main = paste0("cluster: ", cluster))
}

r <- plot_fun("042701010203")
```


apply across all bad raster lists
```{r}
#| eval: false

lapply(bad_list, plot_fun)
```


```{r}
v_r <- values(r[["r"]], na.rm = TRUE)
v_D <- values(r[["DEM"]], na.rm = TRUE)

terra::app
```


The length of the non-NA values is lower for the bad band 

write a new function to capture that and save to a readable dataframe 

```{r}
val_fun <- function(huc){
  l <- list.files(here("Data/HUC_Raster_Stacks/HUC_DL_Stacks/"), full.names = TRUE)
  fn <- l[grepl(huc, l)]
  cluster <- str_extract(fn, "(?<=cluster_)\\d+(?=_)")
  # fn <- paste0("Data/HUC_Raster_Stacks/HUC_DL_Stacks/cluster_", cluster, "_huc_", huc, "_stack.tif")
  r <- rast(here(fn), lyrs = c("DEM", "twi", "CHM", "EVI", "r", "pct_below_1m" ))

  stats <- global(r, fun = "notNA")
  stats$layer <- names(r)
  stats_return <- stats |> arrange(notNA) |> 
    mutate(cluster = cluster, 
           huc = huc)
  return(stats_return)
}

val_l <- lapply(bad_list, val_fun)

```


```{r}
val_df <- bind_rows(val_l)
rownames(val_df) <- NULL
val_df |> group_by(huc) |> mutate(group_mean = mean(notNA),
                                  group_std = sd(notNA),
                                  big_diff = case_when(notNA < abs(group_mean - group_std) ~ "YesBad",
                                                       .default = "No"))
```

```{r}
plot_fun(22, "043002010302")
```

